Trajectories

### 2022-09-07 index.qmd
Report to Quarto

---
title: "Trajectories"
subtitle: "Lab notes"
date: "Updated: `r format(Sys.Date(), '%Y/%m/%d')`"
format:
    html:
        abstract: "Bilinguals face the challenging task of learning words from languages with overlapping phonologies. Floccia et al. (2018) reported larger vocabulary sizes for 24-month-old bilinguals that were learning languages that shared a greater amount of cognates (e.g., English-Dutch). The mechanisms underlying this effect remain unknown. We explore two compatible scenarios. First, we test whether cognates are learnt earlier than non-cognates. This would account for the difference in vocabulary size associated to the amount of shared cognates across languages. Second, we explore the possibility that the word-forms of one language interact with those form the other language, scaffolding the acquisition of their translation equivalents when their phonologies overlap. This mechanism, in line with the parallel activation account of bilingual speech perception, would provide a plausible explanation to why cognates are acquired ealier by bilinguals. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish, and present data on receptive and productive vocabulary of bilingual toddlers aged 12 to 34 months."
        toc: true
        smooth-scroll: true
        link-external-newwindow: true
        number-sections: true
        number-offset: 0
        code-fold: true
        code-overflow: scroll
        code-line-numbers: true
        code-copy: hover
        fig-dpi: 500
        reference-location: margin
        execute-dir: "C:/Users/U155880/Documents/trajectories/"
bibliography: "references.bib"
csl: "apa6.csl"
---

@@

```{r}
#| label: prepare
#| echo: true
#| message: false
#| warning: false
# set params and load target contents
targets::tar_load_globals()
targets::tar_load_everything()

@@

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items (`r n_item_cognate["catalan-cognate"]` cognates, `r n_item_cognate["catalan-noncognate"]` non-cognates) and the Spanish inventory contained `r n_item_language["spanish"]` (`r n_item_cognate["spanish-cognate"]` cognates, `r n_item_cognate["spanish-noncognate"]` non-cognates). Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* [dog] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both in separate items (e.g., Catalan *acabar* [*to finish*] and Spanish *acabar* and *terminar*), or merged them into a single items (e.g., Spanish *mono* [*monkey*] and Catalan *mono/mico*. We included items from a diverse sample of `r n_categories` semantic/functional categories (see Appendix 1). For the analyses included in this study, we excluded items from the adverbs, auxiliary words, connectives, interjections and games and routines categories, so that only data from content words (nouns, adjectives, and verbs) were used.

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Some families filled a long version of the vocabulary checklists (800 translation equivalents; 800 items in Catalan, 800 items in Spanish), while others filled a shorter version (~400 translation equivalents, ~400 items in Catalan, ~400 items in Spanish). These last families were randomly allocated into one of four different subsets of the complete list of items. These lists were carefully designed so that each contained a representative subsample of the items from the complete list. Semantic/functional categories with less than 16 items--thus resulting in less than four items after dividing it in four lists--were not divided in the short version of the questionnaire: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. Table 2 in Appendix 1 shows the distribution of items across questionnaire versions. We excluded from the analysis multi-word items (e.g., *barrita de cereales* [cereal bar]) and items that included more than one word-form (e.g., *mono / mico*). Table 3 shows the classification of items in cognates and non-cognates and their lexical frequency scores across the four lists of the inventories.

@@

## # Phonemes

@@

## Frequency by # Phonemes

@@

## # Phonemes by Levenshtein

@@

@@

:::{.column-body-outset}
$$\begin{aligned}
g[\mathrm{Pr}(y_{ij} \le a_s)] = \tau_s - \nu_{ij}
\end{aligned}$$
:::

@@

:::{.column-body-outset}
$$
\begin{aligned}

\nu_{ij} = & \beta_0 +\beta_1\mathrm{Age}_{i} + \beta_2 \mathrm{Freq}_{j} + \beta_3\mathrm{Phon}_{j} + \beta_4\mathrm{Exposure}_{ij} + \beta_5 \mathrm{Levenshtein}_{j} +
\beta_6\mathrm{(Age_{i} \cdot \mathrm{Exposure}_{ij}}) + \beta_7\mathrm{(\mathrm{Age}_{i} \cdot \mathrm{Levenshtein}_{j}}) +  \beta_8\mathrm{(Exposure_{ij} \cdot \mathrm{Levenshtein}_{j}}) + 
\beta_9{(\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij} \cdot \mathrm{Levenshtein}_{j}}) +\\

@@

&\theta_{0j}^{(\mathrm{ID})} + \theta_{1j}^{(\mathrm{ID})}\mathrm{Age}_{i} +  \theta_{2j}^{(\mathrm{ID})} \mathrm{Freq}_{j} +   \theta_{3j}^{(\mathrm{ID})} \mathrm{Phon}_{j} + \theta_{4j}^{(\mathrm{ID})} \mathrm{Exposure}_{ij} + \theta_{5j}^{(\mathrm{ID})} \mathrm{Levenshtein}_{j} + \theta_{6j}^{(\mathrm{ID})} (\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij}) + 
\theta_{7j}^{(\mathrm{ID})} (\mathrm{Age}_{i} \cdot \mathrm{Levenshtein}_{j}) +
\theta_{8j}^{(\mathrm{ID})} (\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij} \cdot \mathrm{Levenshtein}_{j})

\end{aligned}
$$
:::

@@

:::{.column-body-outset}
$$\begin{aligned}
\beta_{0} & \sim \mathrm{Normal}(-0.25, 0.1) & [\mbox{Intercept/response category threshold}] \\
\beta_{1} &\sim \mathrm{Normal}(1, 0.1) & [\mbox{Age population-level coefficient}]\\
\beta_{2-9} &\sim \mathrm{Normal}(0, 1) & [\mbox{Rest of population-level coefficients}] \\
\theta_{0-9} & \sim \mathrm{Normal}(1, 0.1) & [\mbox{Group-level coefficient variability}]
\end{aligned}
$$
:::

@@

* OS: `r sessionInfo()$running`, `r `sessionInfo()$platform`
* R version: `r sessionInfo()$R.version$version.string`
* Algorithm: Hamiltonian Montecarlo (No U-turn Sampler, *NUTS*)
* Engine: Stan (brms interface, `r sessionInfo()$otherPkgs$brms$Version`)
* brms backend: CmdStanR (cmdstanr `r sessionInfo()$loadedOnly$cmdstanr$Version`)
* Chains: `r dim(model_fit_7$fit)[2]`
* Cores: `r dim(model_fit_7$fit)[2]`
* Iterations: `r dim(model_fit_7$fit)[1]` (`r dim(model_fit_7$fit)[1]/2`)

@@

::: {.panel-tabset}

### Model 0

@@

```{r}
print(model_fit_7$formula)
```

@@

```{r}
#| label: results-raw
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
df %>%
    mutate(
        lv = cut_interval(lv, 5),
        age = floor(age)
    ) %>% 
    count(lv, age, response) %>%
    mutate(prop = n/sum(.$n)) %>% 
    ggplot(aes(age, prop, colour = response, fill = response)) +
    facet_wrap(~lv, nrow = 1) +
    geom_col(position = position_fill()) +
    geom_hline(yintercept = 0.5, colour = "black", size = 0.5) +
    labs(
        x = "Age (months)",
        y = "% responses", 
        colour = "Response",
        fill = "Response"
    ) +
    scale_fill_d3() +
    scale_colour_d3() +
    scale_x_continuous(breaks = seq(0, 40, 2)) +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
    theme(
        legend.position = "top",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()
    )
```

@@

```{r}
#| label: diagnostics-traceplots
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
gather_draws(
    model_fit_7, 
    `b_.*`, `sd_te__.*`, `cor_te__.*`, 
    regex = TRUE, 
    ndraws = 1000
) %>% 
    mutate(.chain = paste("Chain ", .chain)) %>% 
    ggplot(aes(.iteration, .value, colour = .chain)) +
    facet_wrap(~.variable, scales = "free_y") +
    annotate(
        geom = "rect", 
        colour = NA, 
        xmin = 0,
        xmax = dim(model_fit_7$fit)[1]/2, 
        ymin = -Inf, 
        ymax = Inf, 
        alpha = 0.25, 
    ) +
    geom_line() +
    labs(
        x = "Iteration",
        y = "Sample value",
        colour = "Chain"
    ) +
    scale_color_d3() +
    theme(
        legend.position = "top",
        legend.title = element_blank(),
        panel.grid = element_blank()
    )
```

@@

```{r}
#| label: diagnostics-rhat
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
mcmc_rhat_data(rhat(model_fit_7)) %>% 
    ggplot() +
    aes(x = value, fill = description, colour = description) +
    geom_histogram(
        bins = 100, 
        na.rm = TRUE, 
        fill = pal_d3()(1)[1], 
        colour = "white"
    ) +
    geom_rug(alpha = 0.1) +
    labs(
        x = "R-hat",
        y = "Number of samples",
        colour = "Description",
        fill = "Description"
    ) +
    theme(
        legend.position = "top"
    )
```

@@

```{r}
#| label: diagnostics-neff
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
mcmc_neff_data(neff_ratio(model_fit_7)) %>% 
    ggplot() +
    aes(x = value, fill = description, colour = description) +
    geom_histogram(
        bins = 100, 
        na.rm = TRUE, 
        colour = "white"
    ) +
    geom_rug(alpha = 0.01) +
    geom_vline(xintercept = 1, colour = "black", size = 0.75) +
    labs(
        x = "Effective sample size ratio",
        y = "Number of samples",
        colour = "Rating",
        fill = "Rating"
    ) +
    theme(
        legend.position = "top"
    )
```

>>>>

---
title: "Trajectories"
subtitle: "Lab notes"
date: "Updated: `r format(Sys.Date(), '%Y/%m/%d')`"
format:
    html:
        abstract: "Bilinguals face the challenging task of learning words from languages with overlapping phonologies. Floccia et al. (2018) reported larger vocabulary sizes for 24-month-old bilinguals that were learning languages that shared a greater amount of cognates (e.g., English-Dutch). The mechanisms underlying this effect remain unknown. We explore two compatible scenarios. First, we test whether cognates are learnt earlier than non-cognates. This would account for the difference in vocabulary size associated to the amount of shared cognates across languages. Second, we explore the possibility that the word-forms of one language interact with those form the other language, scaffolding the acquisition of their translation equivalents when their phonologies overlap. This mechanism, in line with the parallel activation account of bilingual speech perception, would provide a plausible explanation to why cognates are acquired ealier by bilinguals. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish, and present data on receptive and productive vocabulary of bilingual toddlers aged 12 to 34 months."
        toc: true
        smooth-scroll: true
        link-external-newwindow: true
        number-sections: true
        number-offset: 0
        code-fold: true
        code-overflow: scroll
        code-line-numbers: true
        code-copy: hover
        fig-dpi: 500
        reference-location: margin
        execute-dir: "C:/Users/U155880/Documents/trajectories/"
bibliography: "manuscript/references.bib"
csl: "manuscript/apa6.csl"
---

@@

```{r}
#| label: prepare
#| echo: true
#| message: false
#| warning: false
# set params and load target contents

library(bayesplot)
library(brms) 
library(conflicted)
library(dplyr) 
library(forcats)
library(ggplot2) 
library(ggsci)
library(gt) 
library(here) 
library(janitor)
library(knitr)
library(lubridate)
library(multilex)
library(patchwork) 
library(papaja) 
library(purrr)
library(scales) 
library(stringdist)
library(stringr)
library(targets)
library(tibble)
library(tidybayes) 
library(tidyr)

# load R objects
tar_load_everything()

@@

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items (`r n_item_cognate["catalan-cognate"]` cognates, `r n_item_cognate["catalan-noncognate"]` non-cognates) and the Spanish inventory contained `r n_item_language["spanish"]` (`r n_item_cognate["spanish-cognate"]` cognates, `r n_item_cognate["spanish-noncognate"]` non-cognates). Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* \[dog\] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both in separate items (e.g., Catalan *acabar* \[*to finish*\] and Spanish *acabar* and *terminar*), or merged them into a single items (e.g., Spanish *mono* \[*monkey*\] and Catalan *mono/mico*. We included items from a diverse sample of `r n_categories` semantic/functional categories (see Appendix 1). For the analyses included in this study, we excluded items from the adverbs, auxiliary words, connectives, interjections and games and routines categories, so that only data from content words (nouns, adjectives, and verbs) were used.

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Some families filled a long version of the vocabulary checklists (800 translation equivalents; 800 items in Catalan, 800 items in Spanish), while others filled a shorter version (\~400 translation equivalents, \~400 items in Catalan, \~400 items in Spanish). These last families were randomly allocated into one of four different subsets of the complete list of items. These lists were carefully designed so that each contained a representative subsample of the items from the complete list. Semantic/functional categories with less than 16 items--thus resulting in less than four items after dividing it in four lists--were not divided in the short version of the questionnaire: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. Table 2 in Appendix 1 shows the distribution of items across questionnaire versions. We excluded from the analysis multi-word items (e.g., *barrita de cereales* \[cereal bar\]) and items that included more than one word-form (e.g., *mono / mico*). Table 3 shows the classification of items in cognates and non-cognates and their lexical frequency scores across the four lists of the inventories.

@@

## \# Phonemes

@@

## Frequency by \# Phonemes

@@

## \# Phonemes by Levenshtein

@@

```{r}
#| label: participants-edu-plot
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
participants %>%
    left_join(select(multilex_data$logs, id, time, edu_parent)) %>% 
    mutate(
        doe_2 = pmin(doe_catalan, doe_spanish),
        doe_2 = cut(doe_2, seq(0, 0.5, 0.1), include.lowest = TRUE)
    ) %>% 
    count(doe_2, edu_parent) %>% 
    ggplot() +
    aes(
        x = doe_2,
        y = n,
        fill = edu_parent
    ) +
    geom_col(position = position_fill()) +
    scale_fill_manual(values = sort(pal_d3()(6), decreasing = TRUE)) +
    labs(
        x = "Exposure to second language (L2 DoE)",
        y = "Proportion of participants",
        fill = "Highest parental educational attainment"
    ) +
    scale_y_continuous(labels = percent) +
    theme(
        legend.position = "none",
        panel.grid = element_blank()
    )
```

@@

::: column-body-outset
$$\begin{aligned}
g[\mathrm{Pr}(y_{ij} \le a_s)] = \tau_s - \nu_{ij}
\end{aligned}$$
:::

@@

::: column-body-outset

$$
\begin{aligned}\nu_{ij} = & \beta_0 +\beta_1\mathrm{Age}_{i} + \beta_2 \mathrm{Freq}_{j} + \beta_3\mathrm{Phon}_{j} + \beta_4\mathrm{Exposure}_{ij} + \beta_5 \mathrm{Levenshtein}_{j} +
\beta_6\mathrm{(Age_{i} \cdot \mathrm{Exposure}_{ij}}) + \beta_7\mathrm{(\mathrm{Age}_{i} \cdot \mathrm{Levenshtein}_{j}}) +  \beta_8\mathrm{(Exposure_{ij} \cdot \mathrm{Levenshtein}_{j}}) + 
\beta_9{(\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij} \cdot \mathrm{Levenshtein}_{j}}) +\\

@@

&\theta_{0j}^{(\mathrm{ID})} + \theta_{1j}^{(\mathrm{ID})}\mathrm{Age}_{i} +  \theta_{2j}^{(\mathrm{ID})} \mathrm{Freq}_{j} +   \theta_{3j}^{(\mathrm{ID})} \mathrm{Phon}_{j} + \theta_{4j}^{(\mathrm{ID})} \mathrm{Exposure}_{ij} + \theta_{5j}^{(\mathrm{ID})} \mathrm{Levenshtein}_{j} + \theta_{6j}^{(\mathrm{ID})} (\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij}) + 
\theta_{7j}^{(\mathrm{ID})} (\mathrm{Age}_{i} \cdot \mathrm{Levenshtein}_{j}) +
\theta_{8j}^{(\mathrm{ID})} (\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij} \cdot \mathrm{Levenshtein}_{j})\end{aligned}$$
:::

@@

::: column-body-outset
$$\begin{aligned}\beta_{0} & \sim \mathrm{Normal}(-0.25, 0.1) & [\mbox{Intercept/response category threshold}] \\
\beta_{1} &\sim \mathrm{Normal}(1, 0.1) & [\mbox{Age population-level coefficient}]\\
\beta_{2-9} &\sim \mathrm{Normal}(0, 1) & [\mbox{Rest of population-level coefficients}] \\
\theta_{0-9} & \sim \mathrm{Normal}(1, 0.1) & [\mbox{Group-level coefficient variability}]\end{aligned}$$
:::

@@

-   OS: `r sessionInfo()$running`, `r`sessionInfo()\$platform\`
-   R version: `r sessionInfo()$R.version$version.string`
-   Algorithm: Hamiltonian Montecarlo (No U-turn Sampler, *NUTS*)
-   Engine: Stan (brms interface, `r sessionInfo()$otherPkgs$brms$Version`)
-   brms backend: CmdStanR (cmdstanr `r sessionInfo()$loadedOnly$cmdstanr$Version`)
-   Chains: `r dim(model_fit_7$fit)[2]`
-   Cores: `r dim(model_fit_7$fit)[2]`
-   Iterations: `r dim(model_fit_7$fit)[1]` (`r dim(model_fit_7$fit)[1]/2`)

@@

::: panel-tabset
### Model 0

@@

```{r}
print(model_fit_7$formula)
```
:::

@@

```{r}
#| label: results-raw
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4
df %>%
    mutate(
        lv = cut_interval(lv, 5),
        age = floor(age)
    ) %>% 
    count(lv, age, response) %>%
    mutate(prop = n/sum(.$n)) %>% 
    ggplot(aes(age, prop, colour = response, fill = response)) +
    facet_wrap(~lv, nrow = 1) +
    geom_col(position = position_fill()) +
    geom_hline(yintercept = 0.5, colour = "black", size = 0.5) +
    labs(
        x = "Age (months)",
        y = "% responses", 
        colour = "Response",
        fill = "Response"
    ) +
    scale_fill_d3() +
    scale_colour_d3() +
    scale_x_continuous(breaks = seq(0, 40, 2)) +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
    theme(
        legend.position = "top",
        axis.text.y = element_text(size = 7),
        axis.text.x = element_text(size = 7),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank()
    )
```

@@

```{r}
#| label: diagnostics-traceplots
#| echo: true
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 12
gather_draws(
    model_fit_7, 
    `b_.*`, `sd_te__.*`, `cor_te__.*`, 
    regex = TRUE, 
    ndraws = 1000
) %>% 
    mutate(.chain = paste("Chain ", .chain)) %>% 
    ggplot(aes(.iteration, .value, colour = .chain)) +
    facet_wrap(~.variable, scales = "free_y") +
    annotate(
        geom = "rect", 
        colour = NA, 
        xmin = 0,
        xmax = dim(model_fit_7$fit)[1]/2, 
        ymin = -Inf, 
        ymax = Inf, 
        alpha = 0.25, 
    ) +
    geom_line() +
    labs(
        x = "Iteration",
        y = "Sample value",
        colour = "Chain"
    ) +
    scale_color_d3() +
    theme(
        legend.position = "top",
        legend.title = element_blank(),
        panel.grid = element_blank(),
        axis.text.x = element_text(size = 7)
    )
```

@@

```{r}
#| label: diagnostics-rhat
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| eval: false
mcmc_rhat_data(model_rhats$model_fit_7) %>% 
    ggplot() +
    aes(x = value, fill = description, colour = description) +
    geom_histogram(
        bins = 100, 
        na.rm = TRUE, 
        fill = pal_d3()(1)[1], 
        colour = "white"
    ) +
    geom_rug(alpha = 0.1) +
    labs(
        x = "R-hat",
        y = "Number of samples",
        colour = "Description",
        fill = "Description"
    ) +
    theme(
        legend.position = "top"
    )
```

@@

```{r}
#| label: diagnostics-neff
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| eval: false
mcmc_neff_data(model_neffs$model_fit_7) %>% 
    ggplot() +
    aes(x = value, fill = description, colour = description) +
    geom_histogram(
        bins = 100, 
        na.rm = TRUE, 
        colour = "white"
    ) +
    geom_rug(alpha = 0.01) +
    geom_vline(xintercept = 1, colour = "black", size = 0.75) +
    labs(
        x = "Effective sample size ratio",
        y = "Number of samples",
        colour = "Rating",
        fill = "Rating"
    ) +
    theme(
        legend.position = "top"
    )
```
